%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

\chapter{Preface}

The Preface must include a statement indicating the student's contribution to the following:

\begin{itemize}
  \item Identification and design of the research program,
  \item Performance of the various parts of the research, and
  \item Analysis of the research data.
\end{itemize}

Certain additional elements may also be required, as specified below.

\begin{itemize}
  \item If any of the work presented in the thesis has led to any publications or submissions, all of these must be listed in the Preface. Bibliographic details should include the title of the article and the name of the publisher (ONLY if the article has been accepted or published), and the chapter(s) of the thesis in which the associated work is located.
  \item If the work includes publications or material submitted for publication, the statement described above must detail the relative contributions of all collaborators and co-authors (including supervisors and members of the supervisory committee) and state the proportion of research and writing conducted by the student. For further details, see “Including Published Material in a Thesis or Dissertation”.
  \item If the work includes other scholarly artifacts (such as film and other audio, visual, and graphic representations, and application-oriented documents such as policy briefs, curricula, business plans, computer and web tools, pages, and applications, etc.), all of these must be listed in the Preface (with bibliographical information, if applicable).
  \item If ethics approval was required for the research, the Preface must list the Certificate Number(s) of the Ethics Certificate(s) applicable to the project.
\end{itemize}

The content of the Preface must be verified by the student's supervisor, whose endorsement must appear on the final Thesis/Dissertation Approval form.