\chapter{Acknowledgements}

I'd like to thank the following people for their support over the past two years:

\begin{itemize}
  \item William Bowman, for your encouragement and mentorship throughout my Bachelor's and Master's theses
    and in navigating academia.
    Your influence on me is likely greater than I currently realize,
    and I wouldn't have it any other way.
  \item Alex Summers, for all the helpful feedback you provided as the very second reader of this thesis,
    and for introducing to me a bunch of really fun board games.
  \item Jordy Dickinson, for exposing me to all sorts of niches of type theory and constructive mathematics,
    and for giving an ear to the most out-of-context problems I was working on.
    I hope you're doing well.
  \item Hazel Levine, for making \href{https://types.pl/}{types.pl} a reality (and letting me in on the fun).
  \item The folks on PL Twitter, the TYPES, Coq, Agda, Idris, and Cedille mailing lists and Zulips,
    and the Proof Assistants Stack Exchange for answering my many questions
    and for asking far more interesting ones.
  \item Finally and most importantly, all my other splabmates at the Software Practices Lab
    for making grad school a way better experience than I could have hoped for,
    especially during an ongoing global pandemic.
    To name a few:\footnote{in no particular order except to avoid line breaks across names}
    \vspace{-0.25\baselineskip}
    \begin{quote}
    James Yoo\textsuperscript{\href{https://youtu.be/dQw4w9WgXcQ}{0}}, Chris Chen\punctstack{,}\footnote{honourary splabmate}
    Felipe Ba\~nados Schwerter, Markus de Medeiros, 
    Braxton Hall, Paulette Koronkevich, Justin Frank, Ron Garcia, Lily Bryant,
    Adam Geller, Joey Eremondi,
    \end{quote}
    \vspace{-0.5\baselineskip}
    and many more, especially those who brought treats for the lab.
\end{itemize}

\noindent {\small This research was supported by the Canada Graduate Scholarships -- Master’s (CGS M) programme.
Cette recherche a \'et\'e financ\'ee par le Programme de bourses d'\'etudes sup\'erieures
du Canada au niveau de la maitrise (BESC M).}